tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2024-05-13 response-to-reviewer/response-to-reviewer.tex

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication, and we are extremely excited to hear that this manuscript is acceptable for publication in Bioinformatics subject to the suggested revisions. We are glad that
Reviewers 1 and 2 share our enthusiasm for the tstrait package!
\subsection*{Associate Editor}
% We address your specific comments below:
% % \reviewersection

\textit{
The reviewers are both very positive on your manuscript, rating it highly and seeing no significant problems in need of revision.  Each has a couple of questions for you, and although the reviewers may consider these discretionary I will pass them on to you with a request for a Minor Revision to give you a chance to revise in response to these points before we make a final decision. Given the reviews, I think it is unlikely we will need a second round of review if you can provide reasonable answers to the questions raised.}

\begin{reply}
    Thank you for the helpful feedback. We are grateful for the reviewers 
    comments and have updated the manuscript to clarify.
\end{reply}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
This applications note describes tstrait, which uses ARGs as a basis for simulating traits. This is important because simulated genotypic data use ARGs to quantify what is happening in the genome across many generations.

I thought this paper was very well-written. In particular, Figure 1 is very effective at presenting how traits are simulated. I also thought that the motivation behind the need for an ARG-based trait simulation package was strongly justified. If/when ARGs become more widely used, I think it will be important to have simulation packages like this.
\end{point}

\begin{reply}
    Thank you for the summary. We are delighted to hear your positive feedback.
\end{reply}

\begin{point}
    1) Can dominance and/or epistatic effects be simulated under the framework of, e.g., Figure 1?
\end{point}

\begin{reply}
    Both dominance and epistasis can be simulated under the simulation framework that is described in Figure 1, as it is possible to extract the genetic value of each node at every site. We have included this in the revised manuscript, and we are planning to incorporate these features in the future release.
\end{reply}

\begin{point}
    2) I appreciate the effort to put in the modular framework. With this said, how easy is it to take the simulated traits from tstrait and migrate the files over to R to run, e.g., GWAS and/or genomic prediction?
\end{point}

\begin{reply}
    All simulation results in tstrait are provided as pandas dataframe. Users can simply save the pandas dataframe as a CSV or similar file and load it into other programs. We have added this part in the revised manuscript to emphasize the point.
\end{reply}

\reviewersection

% Point one description
\begin{point}
Tagami et al. present an interesting Python library that can efficiently simulate quantitative traits from ancestral recombination graphs. I like how the authors illustrate the efficiency of this library when considering the size of typical biobank data.
\end{point}

\begin{reply}
    Thank you for your kind comments. We are delighted to hear your positive feedback on this new Python library.
\end{reply}

\begin{point}
    - I was unable to locate the comparisons against AlphaSimR and simplePHENOTYPES.
\end{point}

\begin{reply}
    The comparisons are located in verification.py file in the tstrait GitHub repository 
    to ensure that these statistical tests are done as part of the normal development 
    of the package. As a result, however, they are not very easy to find. We have 
    added a short supplemental note describing these validations, and describing
    where they can be found in the tstrait source code.
\end{reply}

\begin{point}
    Although I appreciate the argument of efficiency, it seems a little exaggerated since for some of those packages (e.g. simplePHENOTYPES), the simulation step only utilizes the QTNs. I.e., although the whole file would have 140 TB, this file would not be needed in its entirety. For a trait with 100 QTNs, only 100 SNPs would be needed for the simulation step.
\end{point}

\begin{reply}
    Although it is correct that methods such as simplePHENOTYPES only use a subset of 
    the genotype matrix, they must still read these large VCFs, which is itself a 
    major undertaking. In practise they read entire VCF into memory.
    For example, simplePHENOTYPES uses the SNPRelate package to first 
    convert the input VCF to GDS format, and then use the \texttt{snpgdsGetGeno}
    function to obtain an explicit genotype matrix in memory for the entire 
    dataset. The sites of interest are then extracted from this large matrix 
    for further analysis. While this could be done more efficiently, 
    any operations on a multi-terabyte VCF are cumbersome and would require
    substantial engineering effort.
\end{reply}

\begin{point}
    Another positive aspect is that the documentation on GitHub is well documented.
\end{point}

\begin{reply}
    Thank you for your feedback, and we agree that documentation is very important.
\end{reply}

\end{document}

### 2024-05-13 cover_letter/cover_letter_resubmission.tex
Revision

I am writing on behalf of my coauthors to submit our manuscript entitled
\emph{tstrait: a quantitative trait simulator for ancestral recombination graphs}
for consideration for publication in Bioinformatics
as an ``Application Note'' research article.

Following recent breakthroughs in simulation and inference,
Ancestral Recombination Graphs (ARGs) are the subject of
intense research interest (see Lewanski et al 2024, PLoS Genetics,
and Brandt et al 2024, GBE, for recent perspectives). One of the key application
areas is statistical genetics, and in particular developing
improved methodology for Genome Wide Association Studies (GWAS).
However, there is currently no available software to efficiently
and conveniently simulate quantitative traits from an ARG,
significantly hampering progress. In this manuscript we describe
\texttt{tstrait}, a new Python package that can efficiently
simulate quantitative traits on ARGs under a wide range of
models. We believe that this tool will be of significant interest
to the readers of Bioinformatics, and will enable further
ground-breaking research in this area.

>>>>

I am writing on behalf of my coauthors to resubmit our manuscript entitled
\emph{tstrait: a quantitative trait simulator for ancestral recombination graphs}
for consideration for publication in Bioinformatics
as an ``Application Note'' research article.

We thank the reviewers for their thoughtful and thorough reviews, and we have addressed those points in the revised manuscript. We have attached
a detailed point-by-point response in the \texttt{response-to-reviewers.pdf}
file.

We have also attached a file showing the changes made to the manuscript
in response to reviewers in \texttt{review-diff.pdf}.

Finally, we attach \texttt{supplementary-material.pdf}, which includes some clarifications on our 
statistical testing methodology.

### 2024-05-13 paper.tex
Revision

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele with $\beta=0.05$,
so all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using standard Python data science tools.

@@

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\section{Supplementary Figures}

\begin{figure}[h]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=\columnwidth]{figures/stdpopsim-time.pdf}
\caption{Time taken to simulate quantitative traits with increasing
sample size. For each sample size, we simulated an ARG under human-like
parameters using the default HomSap demographic model in \texttt{stdpopsim}.
Each point represents the mean time for 10 independent runs of \texttt{tstrait}
for a particular ARG. The times reported are the total CPU
time required to simulate a quantitative trait with 1000 causal sites,
on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
The trait model is a normal distribution with $\mu=0$,
$\sigma^2=1$, $h^2=0.3$, and $\alpha=0$.
}\label{fig:time}
\end{figure}
\end{document}

>>>>

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele with $\beta=0.05$,
so all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
Dominance and epistasis are straightforward extensions to this
model, which we plan to include in future versions of \texttt{tstrait}.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using standard Python data science tools, or exported
to (e.g.) CSV for broad compatibility.

@@

\end{document}
